\LARGE \bf Compressed Directional Distance Transform for Fast 2D Ray Casting

### 2017-02-27 paper/main.tex
several comments

@@

\author{Corey H. Walsh$^{1}$ and Sertac Karaman$^{2}$% <-this % stops a space
% \thanks{*This work was not supported by any organization}% <-this % stops a space
\thanks{$^{1}$Corey H. Walsh is with the Department of Computer Science and Engineering, 
Albert Author is with Faculty of Electrical Engineering, Mathematics and Computer Science, Massachusetts Institute of Technology,
        Cambridge, MA 02139, USA
        {\tt\small chwalsh@mit.edu}}%
\thanks{$^{2}$Sertac Karaman is with Faculty of Computer Science and Engineering, Massachusetts Institute of Technology,
        Cambridge, MA 02139, USA
        {\tt\small sertac@mit.edu}}%
}

@@

@@

@@

To combat this challenge, Fox et al. \cite{localization} suggest the use of a large three-dimensional lookup table (LUT) to store expected ranges for each discrete $(x,y,\theta)$ state. While this is simple to implement and does result in large speed improvements as compared to ray casting, it can be prohibitively memory intensive for large maps and/or resource constrained systems. We propose an analogous data structure called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries.

Many robotic applications require three dimensional maps, however that is considered outside the scope of this work since many data structures exist for accelerating 3D ray casting, such as the bounded volume hierarchy and related variants. The five dimensional table required to precompute expected ranges for a discrete state space $(x,y,z,\theta,\phi)$ would be prohibitively expensive to both compute and store, so 3D acceleration data structures are exclusively used when 3D ray casting is required. Though our approach does extend to higher dimensionality and would be smaller than the analogous five dimensional table, it would still be prohibitively expensive in greater than two dimensions, thus we do not consider it here.

\section{RELATED WORK}

Several widely used algorithms exist for ray casting in a two dimensional space. Due to varied handling of edge cases and ambiguities in the problem of determining distance between any point and the nearest obstacle in a particular direction, most 2D ray casting algorithms do not provide exactly consistent results for every query. However, on average error between results from each of the methods discussed is small.

\subsection{Bresenham's Line}

@@

Ray marching is on average much faster than Bresenham’s line, but edge cases exist in which the performance is equivalent. For example, a query ray which travels closely parallel to a wall will make small steps between each query point for the whole length of the ray. As a result, the performance of ray marching has a long tail distribution which can be problematic for near real time algorithms. Though highly scene dependent, on average ray marching runtime is roughly logarithmic in distance between obstacles, though in the worst case it is linear in map size.

@@

\section{APPROACH}

@@

The euclidean distance transform (EDT) of an occupancy grid map stores the distance to the nearest obstacle in any direction for each point $(x,y)$ in the map. In contrast, a directional distance transform (DDT) stores the distance to the nearest obstacle in a particular direction for each point $(x,y)$ in the map. The key difference between a single Θ slice of the LUT and a directional distance transform is the way in which it is computed. To compute the LUT slice, ray casting is performed in the $\theta_d$ direction from every $(x_d,y_d)$ in the map using another algorithm such as Bresenham’s line. In contrast, to compute the DDT, the obstacles in the map are rotated about the center of the map by $-\theta_d$ and the values of each column in the DDT may be directly computed as a sawtooth function which goes to zero whenever an obstacle is encountered in that column. Thus, each column in the DDT may be characterized as a sawtooth function with uneven periodicity. While it is true that the rotation of the occupancy grid introduces small errors as compared to ray casting for each cell, as previously discussed small errors do not significantly impact particle filter localization performance due to the use of a probabilistic error model. In practice, errors induced by map rotation are dominated by those inherently present in distance sensors.

@@

The conversion from a DDT to a CDDT is performed by converting the sawtooth response of each column into a set of points in which the sawtooth wave goes to zero. This conversion is lossless in the sense that each element of the table may be recovered as a function of the nearest zero point in a particular direction. If the set of zero points is sorted, then the magnitude of the saw tooth wave at any point in the original DDT may be quickly discovered by performing binary search for the nearest zero point in that column and finding the distance between the query point and that zero point. This losslessness and fast indexing allows for ray cast queries to be quickly resolved without storing the entire DDT, but rather by storing the sorted lists of zero points.
For sake of efficiency, the full DDT need not be computed, but rather the CDDT may be built directly by projecting each obstacle into the coordinate space of each theta slice and storing a list of y-coordinates for each discrete x-coordinate. Once all scene geometry is projected into the table, each list of zero points is sorted to facilitate indexing.

@@

@@

In this case, the cost of inserting or deleting an occupied grid cell would be $O(theta\_discretization*\log(number of lut bin elements))$. The number of elements in each lut bin tends to be small and is bounded, so this becomes $O(theta\_discretization)$ which is likely not prohibitive for real-time performance.

@@

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. Consider a 3x3 block of obstacles in an otherwise empty map. The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. This can result in significant memory usage and data structure construction time reductions in dense maps where the number of obstacles adjacent to empty cells is small compared to the total number of obstacles. Prior to performing ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles which may be removed by the morphological operation.

@@

Additionally, consider a line of obstacles aligned along the Y-axis. Every element in this line will be projected into a single list of zero points for the theta=0 slice of the CDDT. However, the middle elements of the line will never result in a collision. Any ray cast from adjacent states in the $\theta=0$ direction will return early in the occupancy grid check, and any ray cast from outside the wall will collide with one of the edge-most obstacles in that line. Therefore in the theta=0 slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary theta not aligned with an axis. Rather than attempting to analytically determine which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state (as in the LUT data structure) and discard any zero point which is unused.

@@

@@

We have implemented the proposed algorithm in C++, as well as Bresenham’s Line, ray marching, and the LUT approach for comparison. Our source code is available for use and analysis [TODO add a link to github], and Python wrappers are also provided for easier usage. All benchmarks were performed on a desktop computer on a Intel Core i5-4590 CPU @ 3.30GHz with 16GB DDR3 ram @ 1333MHz, running Ubuntu 14.04.

>>>>

%: Margin note commands
 \usepackage[usenames,dvipsnames]{xcolor}
\setlength{\marginparwidth}{0.5in}
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%
{\raggedright\footnotesize #1}}
\newcommand{\skcomment}[2]{{\color{red}#1} \marginpar{\color{TealBlue}\tiny #2 (SK)}}

@@

\author{Corey H. Walsh$^{1}$ and Sertac Karaman$^{2}$% <-this % stops a space
% \thanks{*This work was not supported by any organization}% <-this % stops a space
\thanks{$^{1}$Corey H. Walsh is with the Department of Computer Science and Engineering, Massachusetts Institute of Technology,
        Cambridge, MA 02139, USA
        {\tt\small chwalsh@mit.edu}}%
\thanks{$^{2}$Sertac Karaman is with the Laboratory for Information and Decision Systems, Massachusetts Institute of Technology,
        Cambridge, MA 02139, USA
        {\tt\small sertac@mit.edu}}%
}

@@

\skcomment{}{I would consider putting the word ``localization'' in the title.}

@@

\skcomment{}{Perhaps start with a short introduction to localization, and then describe what particle filter (Monte Carlo) localization is, and then introduce ray casting in that context.}

@@

To combat this challenge, \skcomment{Fox et al.}{Thrun et al.?} \cite{localization} suggest the use of a large three-dimensional lookup table (LUT) to store expected ranges for each discrete $(x,y,\theta)$ state. While this is simple to implement and does result in large speed improvements as compared to ray casting, it can be prohibitively memory intensive for large maps and/or resource constrained systems\skcomment{.}{perhaps give an example: In such and such large area with such and such resolution map, the data structure is XX GigaBytes (TeraBytes?)}

We propose an analogous data structure called a Compressed Directional Distance Transform (CDDT) which requires significantly less memory and precomputation time while still allowing near constant time ray casting queries.

\skcomment{}{This is probably a good place to list the contributions of this work as in: ``The contributions of this work are the following. Firstly, we develop ... Secondly, we implement this algorithm for a ... You can also say things like: We observe two orders of magnitude improvement in memory requirements, with little sacrifice in computation time, when compared to lookup table methods. Et cetera.}

\skcomment{Many robotic applications require three dimensional maps, however that is considered outside the scope of this work since many data structures exist for accelerating 3D ray casting, such as the bounded volume hierarchy and related variants.}{I would put the discussion in this paragraph to the future work section in the conclusions, rather than here.} The five dimensional table required to precompute expected ranges for a discrete state space $(x,y,z,\theta,\phi)$ would be prohibitively expensive to both compute and store, so 3D acceleration data structures are exclusively used when 3D ray casting is required. \skcomment{Though our approach does extend to higher dimensionality and would be smaller than the analogous five dimensional table, it would still be prohibitively expensive in greater than two dimensions, thus we do not consider it here.}{I would not worry about putting this justification either.}

\skcomment{
\section{RELATED WORK}
}{You might want to integrate this section into the introduction section. Otherwise the introduciton section is unusually short. Many people discuss related work in the introduciton. You can put this discussion right after your second paragraph, and right before listing all your contributions.}
Several widely used algorithms exist for ray casting in a two dimensional space. Due to varied handling of edge cases and ambiguities in the problem of determining distance between any point and the nearest obstacle in a particular direction, most 2D ray casting algorithms do not provide exactly consistent results for every query. However, on average error between results from each of the methods discussed is small.

\skcomment{\subsection{Bresenham's Line}}{I would not worry about these section headings. You say the names of the algorithms in the beginning of the paragraphs anyways.}

@@

Ray marching is on average much faster than Bresenham’s line, but edge cases exist in which the performance is equivalent. For example, a query ray which travels closely parallel to a wall will make small steps between each query point for the whole length of the ray. As a result, the performance of ray marching \skcomment{has a long tail distribution}{is this widely known? Is there a paper that you can cite here? Or can you refer the reader to your results in this paper (e.g., write ``(See Section X.)'' to tell the reader that there are some results there on heavy tails of this algorithm.} which can be problematic for near real time algorithms. Though highly scene dependent, on average ray marching runtime is roughly logarithmic in distance between obstacles, though in the worst case it is linear in map size.

@@

\skcomment{
\section{APPROACH}
}{Instead of this section title, i would use ``The Compressed Directional Distance Transform (CDDT) Algorithm'' or something like that.}

@@

The euclidean distance transform (EDT) of an occupancy grid map stores the distance to the nearest obstacle in any direction for each point $(x,y)$ in the map. In contrast, a directional distance transform (DDT) stores the distance to the nearest obstacle in a particular direction for each point $(x,y)$ in the map. The key difference between a single Θ slice of the LUT and a directional distance transform is the way in which it is computed. To compute the LUT slice, ray casting is performed in the $\theta_d$ direction from every $(x_d,y_d)$ in the map using another algorithm such as Bresenham’s line. In contrast, to compute the DDT, the obstacles in the map are rotated about the center of the map by $-\theta_d$ and the values of each column in the DDT may be directly computed as a sawtooth function which goes to zero whenever an obstacle is encountered in that column. \skcomment{}{Do you think this is a good place for an example?} Thus, each column in the DDT may be characterized as a sawtooth function with \skcomment{uneven periodicity}{what is uneven periodicity mean?}. While it is true that the rotation of the occupancy grid introduces small errors as compared to ray casting for each cell, as previously discussed small errors do not significantly impact particle filter localization performance due to the use of a probabilistic error model. In practice, errors induced by map rotation are dominated by those inherently present in distance sensors.

@@

\skcomment{The conversion from a DDT to a CDDT is performed by converting the sawtooth response of each column into a set of points in which the sawtooth wave goes to zero.}{This is probably the most important part, but kind of hard to understand. I would consider an example here as well.} This conversion is lossless in the sense that each element of the table may be recovered as a function of the nearest zero point in a particular direction. If the set of zero points is sorted, then the magnitude of the saw tooth wave at any point in the original DDT may be quickly discovered by performing binary search for the nearest zero point in that column and finding the distance between the query point and that zero point. This losslessness and fast indexing allows for ray cast queries to be quickly resolved without storing the entire DDT, but rather by storing the sorted lists of zero points.
For sake of efficiency, the full DDT need not be computed, but rather the CDDT may be built directly by projecting each obstacle into the coordinate space of each theta slice and storing a list of y-coordinates for each discrete x-coordinate. Once all scene geometry is projected into the table, each list of zero points is sorted to facilitate indexing.

\skcomment{}{This section could be organized a bit better. Perhaps have a subsection that says ``Compression algorithm" (text above here) and another one that says ``Query Algorithm'' (text below here).}

@@

\skcomment{}{You can put all the text after this point into one subsection, which you can call ``Further optimizations'' or something like that.}

@@

In this case, the cost of inserting or deleting an occupied grid cell would be \skcomment{$O(theta\_discretization*\log(number of lut bin elements))$.}{I think this is your plan already, but just to note: Please make sure to use symbols for these and introduce what the symbols mean. } The number of elements in each lut bin tends to be small and is bounded, so this becomes $O(theta\_discretization)$ which is likely not prohibitive for real-time performance.

@@

By removing entries in the CDDT which can never possibly result in a ray collision, it is possible to further reduce the memory footprint of the data structure. \skcomment{Consider a 3x3 block of obstacles in an otherwise empty map.}{I think this deserves a figure, just to allow the reader to quickly read your paper.} The center obstacle will never be the nearest neighbor in any ray casting query, because any such query would first intersect with one of the 8 surrounding obstacles. To exploit this, one can take the morphological edge map of the occupancy grid map prior to CDDT generation without loss of generality. This can result in significant memory usage and data structure construction time reductions in dense maps where the number of obstacles adjacent to empty cells is small compared to the total number of obstacles. Prior to performing ray casting on the CDDT data structure, we check the occupancy grid to ensure that the query point itself is not an obstacle and return a distance of zero if so in order to prevent incorrect results when ray casting from the middle of obstacles which may be removed by the morphological operation.

@@

\skcomment{Additionally, consider a line of obstacles aligned along the Y-axis.}{While you are at it, put a figure for this one too? } Every element in this line will be projected into a single list of zero points for the theta=0 slice of the CDDT. However, the middle elements of the line will never result in a collision. Any ray cast from adjacent states in the $\theta=0$ direction will return early in the occupancy grid check, and any ray cast from outside the wall will collide with one of the edge-most obstacles in that line. Therefore in the theta=0 slice, one may discard the zero points corresponding to the middle elements without introducing error. This form of optimization is simple to compute in the cardinal directions, but non-trivial for arbitrary theta not aligned with an axis. Rather than attempting to analytically determine which obstacles may be discarded, it is simpler to prune the data structure by ray casting from every possible state (as in the LUT data structure) and discard any zero point which is unused.

@@

\skcomment{}{Do you think you can have a subsection called ``analysis'' which has a few propositions/theorems that say things like Memory requirement is $O(NM)$ and running time is $O(1)$? This may be hard given the deadline, but osmething to think about. }

@@

We have implemented the proposed algorithm in the C++ programming language, as well as Bresenham’s Line, ray marching, and the LUT approach for comparison. Our source code is available for use and analysis [TODO add a link to github], and Python wrappers are also provided for easier usage. All benchmarks were performed on a desktop computer on a Intel Core i5-4590 CPU @ 3.30GHz with 16GB DDR3 ram @ 1333MHz, running Ubuntu 14.04.
